\documentclass[11pt]{article}
\usepackage[margin=2.6cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{parskip}
\usepackage[T1]{fontenc}
\usepackage{booktabs}

\newcommand{\Wt}{\widetilde W}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\bra}[1]{\langle #1|}

\title{Why a Markov chain is secretly an imaginary-time quantum system\\
\large The Rokhsar--Kivelson construction, and how the program's measurements slot into it}
\author{}
\date{2026--07--16}

\begin{document}
\maketitle

\section{Setup}

Let $\Omega$ be the configuration space (triangulations of $S^3$ modulo isomorphism ---
countable, so everything below is honest linear algebra), let
$\pi(T) \propto e^{-S(T)}$ be the target Gibbs measure, and let
$W(T' \!\leftarrow\! T)$ be the Metropolis transition matrix over Pachner moves. The one
property engineered into $W$ is \textbf{detailed balance}:
\[
\pi(T)\, W(T'\!\leftarrow\!T) \;=\; \pi(T')\, W(T\!\leftarrow\!T') .
\]

\section{The symmetrization trick}

$W$ is not symmetric, but detailed balance says it is \emph{conjugate} to a symmetric
matrix. Let $D = \operatorname{diag}(\pi)$ and define
\[
\Wt \;=\; D^{-1/2}\, W\, D^{1/2},
\qquad\text{i.e.}\qquad
\Wt(T',T) \;=\; \pi(T')^{-1/2}\; W(T'\!\leftarrow\!T)\; \pi(T)^{1/2} .
\]
Detailed balance is \emph{exactly} the statement $\Wt^{\dagger} = \Wt$. Define
\[
\boxed{\;H \;=\; \mathbf{1} - \Wt\;}
\]
a self-adjoint operator on $\ell^2(\Omega)$ with three properties, each verifiable in a
line:
\begin{enumerate}
\item \textbf{Ground state.} The vector $\psi_0(T) = \sqrt{\pi(T)}$ satisfies
$\Wt\,\psi_0 = \psi_0$ (unwind the conjugation: this is just stationarity
$W\pi = \pi$). Hence $H\psi_0 = 0$.
\item \textbf{Positivity.} All eigenvalues of $\Wt$ lie in $[-1,1]$ (it is a
similarity transform of a stochastic matrix), so $H \succeq 0$: $\psi_0$ is the
\emph{ground} state, at energy exactly zero.
\item \textbf{Uniqueness.} If the chain is irreducible (ergodic), Perron--Frobenius
makes $\psi_0$ the unique ground state --- and $\psi_0 > 0$ everywhere, a
\emph{nodeless} wavefunction. Hamiltonians whose ground states can be written nodeless
in a preferred basis are called \emph{stoquastic}: no sign problem.
\end{enumerate}

\section{The dictionary}

Run the chain for $t$ steps: probabilities evolve by $W^t$. Conjugating, amplitudes
evolve by
\[
\Wt^{\,t} \;=\; (\mathbf{1} - H)^t \;\approx\; e^{-Ht},
\]
which is \textbf{imaginary-time Schr\"odinger evolution}, $\partial_t \psi = -H\psi$ ---
not an analogy but the same matrix exponentiated. The consequences follow mechanically.

\paragraph{Equal-time expectations are ground-state expectations.}
For a diagonal observable $\hat O$ (a function of the configuration),
\[
\bra{\psi_0}\hat O\ket{\psi_0} \;=\; \sum_T \pi(T)\, O(T),
\qquad\text{since } \psi_0^2 = \pi .
\]
The Born rule \emph{is} the Gibbs measure. This is the precise sense of ``the ensemble
is $|\Psi|^2$.''

\paragraph{Autocorrelation times are inverse energy gaps.}
Expanding a connected time-correlation function in eigenstates
$H\psi_n = E_n\psi_n$:
\[
\big\langle \delta O(0)\,\delta O(t) \big\rangle_{\text{chain}}
\;=\; \sum_{n\ge 1} \big|\bra{\psi_n}\hat O\ket{\psi_0}\big|^2\, e^{-E_n t}.
\]
The slowest decay is $e^{-E_1 t}$: the measured $\tau$'s of the convergence gate are
literally $1/E_n$, and the GEVP analysis is textbook variational spectroscopy of $H$
--- which is why lattice-QCD machinery (L\"uscher--Wolff) applied without modification.
Every $\widehat R$ certificate ever produced was, unknowingly, an energy-gap
measurement.

\section{What is canonical and what is not}

The ground state $\psi_0 = \sqrt\pi = e^{-S/2}$ depends \textbf{only on the action}
$S$. But $H$ itself depends on the \emph{move set and proposal rates}: different
ergodic kernels for the same $\pi$ give different excited spectra above the same ground
state. This single fact is simultaneously:
\begin{itemize}
\item the \textbf{non-uniqueness problem} (which $H$ is ``physical''? --- none
canonically);
\item the \textbf{lapse freedom} (the implementable family is
$H[N] = \sum_x N_x h_x$ with each local $h_x \succeq 0$ annihilating $\psi_0$; uniform
sweeps are the unit-lapse gauge);
\item a \textbf{protection clause}: any measurement touching only the measure ---
roundness, structure factors, hyperuniformity --- is move-set-independent, hence robust
physics of the \emph{state}; anything dynamical (gaps, dynamical exponents,
heterogeneity) is physics of the state \emph{plus a choice of clock}.
\end{itemize}

\section{Where ``RK'' enters}

Rokhsar and Kivelson (1988), studying quantum dimer models, ran the construction
\emph{backwards}: they engineered a quantum Hamiltonian whose ground state is the
equal-weight superposition of dimer configurations --- choosing a classical measure and
building its parent Hamiltonian. The general lesson (Henley 2004; Castelnovo et al.\
2005) is that every reversible chain sits at such an ``RK point'': a locus in
Hamiltonian space where the ground state is exactly a classical Gibbs measure in
disguise. RK phenomenology then supplies sharp expectations --- most famously, when the
classical measure is \emph{critical}, the parent Hamiltonian is gapless with dynamical
exponent $z = 2$ (diffusive: $E_1 \sim k^2$, so gap $\sim L^{-2}$).

\section{How the program's measurements slot in}

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
phase & measured gap & RK reading \\
\midrule
extended ($\beta>0$) & $E_1 \propto N^{0}$ (gapped) &
\begin{tabular}[t]{@{}l@{}}classical measure non-critical (plain fluid)\\
$\Rightarrow$ parent Hamiltonian boringly gapped\end{tabular} \\[1.4em]
scale-free hub ($\beta=0$, flat pin) & $E_1 \propto N^{-0.38}$ &
\begin{tabular}[t]{@{}l@{}}diverging classical correlation structure\\
(hub hierarchy) $\Rightarrow$ gap closes with the\\
\emph{same} exponent: RK phenomenology in the wild\end{tabular} \\[2.2em]
crumpled ($\beta=0$, $k=0$) & $E_1 \propto N^{-0.19}$ & expander/mean-field-like \\
\bottomrule
\end{tabular}
\end{center}

The gapped extended phase explains why the flagship $S(k\to 0)$ test came out thermal,
and why gravity-like structure --- which needs \emph{extensive} zero modes, an actual
constraint surface --- is absent: the kernel of our $H$ is the single ray
$\mathbb{R}\,\psi_0$.

Connecting to the ADM primer: each local generator $h_x$ annihilates $\psi_0$
\emph{separately} (detailed balance holds move-by-move), so ``constraint propagation''
is free --- and vacuous. The constraint structure is trivially abelian. Gravity's
content lives in constraints that close only \emph{jointly}, with field-dependent
structure functions (the Dirac hypersurface-deformation algebra). That gap --- trivial
algebra versus Dirac algebra --- is the distance between ``a sampler with a lapse'' and
``gravity.''

\section{The Euclidean caveat}

The construction is intrinsically \textbf{Euclidean}: $e^{-Ht}$ is a heat semigroup;
nothing here produces a unitary $e^{-iHt}$ or a preferred analytic continuation, and
the chain's $t$ is the unit-lapse gauge clock, not a physical time. Everything
real-time or relational (matter clocks, Page--Wootters) is a further construction,
deliberately deferred.

\paragraph{References.}
D.~Rokhsar, S.~Kivelson, Phys.\ Rev.\ Lett.\ \textbf{61}, 2376 (1988).
C.~Henley, J.~Phys.: Condens.\ Matter \textbf{16}, S891 (2004) --- the clearest
exposition of the chain--Hamiltonian dictionary.
C.~Castelnovo, C.~Chamon, C.~Mudry, P.~Pujol, Ann.\ Phys.\ \textbf{318}, 316 (2005) ---
the general stochastic-matrix-form construction.
M.~L\"uscher, U.~Wolff, Nucl.\ Phys.\ B \textbf{339}, 222 (1990) --- the GEVP method.

\end{document}
