\documentclass[11pt]{article}
\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black]{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem*{remark}{Remark}

\newcommand{\kh}{\hat{\mathbf k}}
\newcommand{\kv}{\mathbf k}
\newcommand{\nh}{\hat{\mathbf n}}
\newcommand{\Nfk}{n_6}
\DeclareMathOperator{\lk}{lk}
\DeclareMathOperator{\divg}{div}

\title{A Minimal Action for Triangulated Gravity Statics\\
and the Emergent Gauge Structure of the Six-Edge Disclination Web\\[4pt]
\large Session notes: aesthetics of the objective function, the fullerene theorem,\\
the $S_{66}(\mathbf k)$ pinch-point measurement, and the exact-segment vector channel}
\author{Working notes, Discrete Differential Geometry project}
\date{July 20--22, 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
We summarize a sequence of results on the statistical mechanics of $3$-torus
triangulations sampled by Metropolis--Hastings over bistellar and hinge moves,
viewed as a discrete ADM-type model: a tetrahedrally close-packed (TCP)
crystalline host carrying a gas of quantized curvature defects. The main
findings: (i) the working objective decomposes into a cosmological volume pin,
an $R^2$-type curvature-square term on hinges, and a Frank--Kasper legality
potential; (ii) an ablation ladder shows the third term is \emph{almost}
redundant --- above a sharp stiffness threshold the two-term action alone
sustains the crystal vacuum and a quantized defect gas; (iii) the remaining
content of the legality term splits into a part that is a \emph{theorem}
(disclination lines cannot end at edge-legal vertices, by the nonexistence of
one-hexagon fullerenes) and a part that is a weak junction-valence cap;
(iv) the degree-six edge web behaves as an emergent gauge field whose lattice
Gauss law we measured through the longitudinal suppression of its director
structure factor $S_{66}(\mathbf k)$, screened by monopoles (legality
violations) at the observed density; and (v) an exact-segment-integral
construction upgrades this diagnostic from a smeared inequality to an identity:
the longitudinal power of the oriented six-flux equals the monopole-charge
structure factor divided by $k^2$, vanishing identically on a closed web.
Supporting results on host-crystal choice, constraint-preserving block moves,
a frozen-subset sampler, and equilibrium certification are collected along the
way.
\end{abstract}

\tableofcontents

%--------------------------------------------------------------------
\section{Setting and notation}
\label{sec:setting}

Configurations are combinatorial triangulations of $T^3$ with $N_3$
tetrahedra, all simplices equilateral in the Regge sense. Curvature is
concentrated on hinges (edges): an edge $e$ of degree $\deg e$ carries deficit
\begin{equation}
\delta_e \;=\; 2\pi - \theta\,\deg e,
\qquad
\theta = \arccos\tfrac13 \approx 1.230959\,,
\label{eq:deficit}
\end{equation}
so the degree of an exactly flat hinge is $2\pi/\theta \approx 5.104299$,
and the physical degrees carry
\[
\delta_3 = +2.590,\quad
\delta_4 = +1.359,\quad
\delta_5 = +0.128,\quad
\delta_6 = -1.103 \quad (\mathrm{rad}).
\]
Since flat degree is not an integer, flat space is geometrically frustrated:
the ground states are tetrahedrally close-packed (Frank--Kasper) structures
\cite{frank-kasper,nelson,sadoc-mosseri} whose edges are degree $5$ and $6$
only, with the degree-6 edges forming a closed line network --- the
\emph{disclination web} (each six-edge is a $-72^\circ$ disclination
segment). Vertex classes are labelled by $\Nfk$, the number of incident
six-edges: $Z12$ ($\Nfk=0$), $Z14$ ($\Nfk=2$, line passes through), $Z15$
($\Nfk=3$, Y-junction), $Z16$ ($\Nfk=4$, X-junction). \emph{FK-legality}
comprises the edge condition ($\deg e \in \{5,6\}$) and the class condition
($\Nfk \in \{0,2,3,4\}$).

The sampler is Metropolis over bistellar (Pachner) moves plus hinge moves,
with a configurable objective. The production host is the TCP crystal $R$
in an $m\times m\times m$ supercell ($m=4$ throughout: $N_3 \approx 47{,}500$,
$E \approx 68{,}000$ edges, six-fraction $\approx 0.105$). Defect positions
use harmonic torus (cocycle) coordinates maintained incrementally by the
sampler. The elementary excitation of the crystal is the
\emph{$(3,4,4)$-knot}: an exact positive-disclination knot of one
degree-3 and two degree-4 edges, carrying quantized curvature
$q \approx -1.05$ rad relative to the cell mean. Equilibrium claims are
gated by the project standard: quantized split-$\hat R < 1.05$, minimum
effective sample size $\geq 100$, two-sided initialization.

%--------------------------------------------------------------------
\section{What the working objective secretly is}
\label{sec:decomposition}

The question posed: \emph{what is the most aesthetically pleasing objective
function --- what most naturally captures local GR dynamics?} The original
motivation was the picture of a gravitational wave distorting a sphere
without changing its volume (hence the early vertex-degree-variance
penalties). Decomposing the objective that actually produced the defect
liquid ($\lambda = 0.4$):

\begin{center}
\begin{tabular}{lll}
\toprule
term & implementation & gravity reading \\
\midrule
volume pin & $(N_3-\bar N)^2$ & cosmological-constant / lapse-fixing term \\
EDQ & $\sum_e (\deg_e - \bar d)^2 \propto \sum_e(\delta_e-\bar\delta)^2$ &
$R^2$ curvature-square on hinges \cite{hamber-williams} \\
$n_6$ potential & $\mathrm{dist}^2(\Nfk,\{0,2,3,4\}) + m^2$ &
FK scaffolding (the inelegant part) \\
\bottomrule
\end{tabular}
\end{center}

Two of the three terms are standard gravity actions in disguise. Moreover,
for equilateral complexes the Regge (Einstein--Hilbert) action is
\emph{linear} in the $f$-vector,
\begin{equation}
\sum_e \delta_e \;=\; 2\pi E - 6\,\theta N_3\,,
\end{equation}
so ``EH $+$ $\Lambda$'' is precisely the classic dynamical-triangulations
action \cite{ajl}, of which our pins are the soft Lagrange-multiplier
version; the mean-degree ``bump'' knob is literally a background-curvature
source.

\paragraph{What was right and wrong about the volume-preserving instinct.}
The instinct --- gap the scalar/volume sector, leave the transverse-traceless
sector soft --- targets exactly the channel structure later measured in the
liquid (scalar power $0.0023\pm0.0004$ vs.\ TT/V/L $\approx 0.011$, a
$5\times$ gap). The error was in the variable: vertex degree is a proxy for
local volume $\sqrt g$, but in this geometry \emph{flat space is not
vertex-homogeneous} --- the TCP vacuum requires coordinations
$12/14/15/16$ in crystalline proportion, so a vertex-degree-variance penalty
declares the vacuum itself defective and fights the crystal. In Regge
calculus curvature lives on hinges, not vertices; the corrected form of the
same instinct is: fix the volume \emph{globally} (one soft $\Lambda$ term
--- total volume, which a passing wave preserves), penalize \emph{curvature}
locally ($R^2$ on hinges). Volume-preserving shape distortions ---
rearrangements of the six-web at fixed deficit budget --- remain the soft
directions. This suggests the candidate minimal action
\begin{equation}
\boxed{\;S \;=\; c_\Lambda\,(N_3-\bar N)^2 \;+\;
c_R \sum_e (\delta_e - \bar\delta)^2\;}
\label{eq:twoterm}
\end{equation}
--- two couplings, two geometric invariants, no privileged vertex
taxonomies --- \emph{if} the $n_6$ term proves redundant.

\paragraph{Two sharpenings before the test.}
(1) \emph{EDQ is blind within the pure-$\{5,6\}$ manifold.} The deficits are
a function of the degree multiset alone, so every pure-$\{5,6\}$ state at a
given six-fraction has identical EDQ energy: whatever order emerges under
EDQ-only is selected by entropy and realizability, not energy. Note also
the link sum rule: for a vertex whose link is a triangulated $S^2$ with
vertex valences equal to the incident edge degrees, Euler's formula gives
$\sum_{v \in \lk} (6 - \mathrm{val}(v)) = 12$; hence pure-$\{5,6\}$ forces
\emph{exactly twelve} five-edges per vertex. Edge-legality alone already
fixes most of the FK vertex structure. What it does not obviously force is
the class condition: $\Nfk = 1$ (``Z13'') and $\Nfk \geq 5$ (``Z17$^+$'')
are edge-legal but class-illegal.
(2) \emph{The $n_6$ term, decoded, is a line-closure condition.} Reading
$\{0,2,3,4\}$ as web topology (none / passthrough / Y / X), the forbidden
$\Nfk = 1$ is a disclination line \emph{ending in the bulk} --- a six-web
monopole. ``Disclination lines do not end'' is a genuine identity of
continuum disclination theory (a discrete Bianchi-type condition), so the
term is respectable physics in an ugly implementation. The ablation question
therefore sharpens to: \emph{is line closure emergent, or must it be
imposed?}

%--------------------------------------------------------------------
\section{The EDQ-only ladder: a sharp threshold}
\label{sec:ladder}

The test: rerun the coupling ladder with the $n_6$ potential deleted
($z_{\rm leg} = c_{\rm imp} = 0$), EDQ and volume pin retained, $m=4$ host,
$12$k sweeps per rung.

\begin{center}
\begin{tabular}{lllll}
\toprule
$\lambda_{\rm EDQ}$ & state & legal-edge frac & legal-vertex frac & $\langle \deg_e\rangle$ \\
\midrule
0.4 & crumple-melted & 0.60 & 0.0002 & \textbf{5.62} \\
0.8 & dense disclination fluid & 0.80 & 0.057 & 5.115 \\
1.2 & \textbf{crystal + defect gas} & --- & 0.991 & 5.10432 \\
1.6 & crystal, $n_{\rm ill}=2$ & --- & $\approx 1$ & flat \\
2.4 & perfect crystal & 1.0 & 1.0 & flat \\
\bottomrule
\end{tabular}
\end{center}

Two failure modes at low stiffness are physically instructive. At
$\lambda_{\rm EDQ}=0.4$ (the stiffness at which the three-term liquid is
stable) the mean degree runs away to $5.62$: with only a soft quadratic
regulator the state crumples --- the classic dynamical-triangulations
crumpling instability leaking through an undersized $R^2$ term. At $0.8$
the mean degree is held but pure-$\{5,6\}$ order does not assemble
(legal-vertex fraction $6\%$), consistent with EDQ's energetic blindness on
the constraint surface. Above a sharp threshold between $0.8$ and $1.2$,
however, the two-term action \eqref{eq:twoterm} sustains \emph{both} phases
of interest: the TCP crystal vacuum ($\lambda_{\rm EDQ} \gtrsim 1.6$) and a
crystal-hosted quantized defect gas ($\lambda_{\rm EDQ} \approx 1.2$,
$n_{\rm ill} \sim 60$--$90$, legal-vertex fraction $0.991$, mean degree
pinned at flat). (Caveat: the $\lambda_{\rm EDQ}=1.2$ gas showed a residual
drift $61 \to 92$ over the $12$k-sweep window and has not yet passed the
certification gates; threshold-region sensitivity is expected.)

%--------------------------------------------------------------------
\section{The census and the fullerene theorem}
\label{sec:fullerene}

If line closure must be imposed, the EDQ-only gas should swarm with Z13
endpoints. The census over pure-$\{5,6\}$ vertices:

\begin{center}
\begin{tabular}{lccc}
\toprule
state & edge-illegal vertices & Z13 endpoints ($\Nfk{=}1$) & Z17$^+$ ($\Nfk\geq 5$) \\
\midrule
$\lambda_{\rm EDQ}=1.2$ gas (no $n_6$ term) & 102 & \textbf{0} & 21 \;{\footnotesize($\Nfk = 5{:}14,\; 6{:}5,\; 8{:}2$)} \\
$\lambda_{\rm EDQ}=1.6$ cold (no $n_6$ term) & 0 & \textbf{0} & 0 \\
$\lambda=0.4$ liquid (with $n_6$ term) & 61 & \textbf{0} & 1 \\
\bottomrule
\end{tabular}
\end{center}

Zero line endpoints appear anywhere --- even with the closure term deleted.
This is not luck; it is a theorem.

\begin{theorem}[line closure is free]
No vertex of a combinatorial $3$-manifold whose incident edges all have
degree $5$ or $6$ can have exactly one incident degree-6 edge.
\end{theorem}

\begin{proof}[Proof sketch]
The link of such a vertex is a triangulated $S^2$ whose vertex valences are
the incident edge degrees: twelve $5$-valent vertices (link sum rule,
\S\ref{sec:decomposition}) and $h$ $6$-valent ones, where $h = \Nfk$. Its
dual is a trivalent polyhedron with twelve pentagonal and $h$ hexagonal
faces --- a fullerene. Fullerenes exist for every $h \geq 0$ \emph{except}
$h = 1$ \cite{grunbaum-motzkin}. Hence $\Nfk = 1$ is combinatorially
impossible at an edge-legal vertex.
\end{proof}

So the ``discrete Bianchi identity'' half of the $n_6$ term was enforcing a
combinatorial impossibility: \emph{web closure follows from edge-legality
alone}. The term's only non-redundant work is suppressing Z17$^+$
over-branchings --- present at $21/10{,}176 \approx 0.2\%$ of vertices in
the EDQ-only gas, and arguably physically admissible anyway (higher-valence
junctions occur in real TCP and quasicrystalline systems).

\paragraph{The aesthetic verdict.}
The two-term action \eqref{eq:twoterm} generates the whole hierarchy:
\[
R^2\ \text{stiffness}
\;\Rightarrow\; \text{pure-}\{5,6\}\ \text{preference}
\;\Rightarrow\; \underbrace{\text{closed web}}_{\text{fullerene theorem}}
\;\Rightarrow\; \text{emergent Gauss law}
\;\Rightarrow\; \text{(with monopole suppression) Coulomb-phase statics.}
\]
FK-legality, line closure, and the gauge structure are theorems or
thermodynamics --- none of them axioms. The original ``volume fixed, shape
free'' instinct survives exactly, rephrased on hinges. Each arrow above now
has either a measurement or a proof behind it; the last one
(\S\ref{sec:s66}--\ref{sec:vector}) is measured but currently
monopole-screened.

%--------------------------------------------------------------------
\section{Interlude: is $R$ the right host crystal?}
\label{sec:host}

\begin{center}
\begin{tabular}{lllll}
\toprule
host & native $\langle\deg\rangle$ & offset from flat ($5.10430$) & six-web & classes \\
\midrule
$R$ & 5.10423 & $-7\times10^{-5}$ (flat) & bent lines, Y and X junctions & all four \\
C15 (Laves) & 5.1000 & $-0.0043$ & diamond lattice of X-nodes & Z12, Z16 \\
A15 & 5.1111 & $+0.0068$ & parallel straight lines, no junctions & Z12, Z14 \\
$\sigma$ & 5.1089 & $+0.0046$ & planar nets & Z12/14/15 \\
\bottomrule
\end{tabular}
\end{center}

$R$'s native mean degree sits within $7\times10^{-5}$ of the ideal flat
value --- as far as we know \emph{the} flat TCP in the equilateral-Regge
sense. This is load-bearing for the charged-gas program: with $R$,
``unpinned vacuum'' $\approx$ ``zero-curvature slice,'' so the bump knob
dopes charge around genuine neutrality. C15 sits on a permanent $-0.0043$
offset (pinning it flat is already a hard positive bump), so the
dilute-neutral regime does not exist there. C15's roles are elsewhere:
host-independence control for the channel measurements, the natural
laboratory for the block-move/transport program (the synchro-Shockley /
polytype literature lives in the Laves family, and its web --- a diamond
lattice of X-junctions --- is the simplest closed network), and a neutral
homogeneous positive-curvature (de Sitter-like) background for free. A15 is
the ``hydrogen atom'' of line dynamics: three families of straight,
junction-free disclination lines. Conclusion: keep $R$ primary for neutral
statics; migrate interpretive web work to C15/A15 when the six-web becomes
the protagonist.

%--------------------------------------------------------------------
\section{The block-move algebra (status: aspiration with three pillars)}
\label{sec:blockmoves}

``Block move algebra'' is our coinage, naming a compound of established
results plus one project-specific aspiration.

\emph{Pillar 1 --- moves as dynamics in discrete gravity.} Pachner's theorem
\cite{pachner} (bistellar moves connect PL-homeomorphic triangulations) is
the ergodicity statement for the unconstrained manifold; Sorkin's tent moves
\cite{sorkin} and Dittrich--H\"ohn's canonical simplicial gravity
\cite{dittrich-hohn} formulate discrete time evolution literally as Pachner
move sequences, with the Bahr--Dittrich warning that discretization
generically breaks diffeomorphism symmetry \cite{bahr-dittrich}. In LQG the
Hamiltonian constraint acts by local graph moves \cite{thiemann}; in TQFT
the $2\!\leftrightarrow\!3$ move consistency \emph{is} the pentagon identity
--- the cleanest precedent for ``relations among moves are algebraic
identities.''

\emph{Pillar 2 --- move sets on constrained manifolds.} Worm and loop
algorithms \cite{prokofev-svistunov}, and quasicrystal phason flips, all
solve our structural problem: single local moves exit the constraint
manifold, so one builds composite moves that stay on it and proves
flip-connectivity. Single Pachner moves provably cannot preserve
FK-legality; the constrained analog of Pachner's theorem
(``FK-flip-connectivity'') is the hard open theorem.

\emph{Pillar 3 --- the target structure.} The hypersurface-deformation
algebra \cite{adm}
\begin{equation}
\{H[N], H[M]\} \;=\; D\!\left[q^{ab}(N\partial_b M - M\partial_b N)\right]:
\end{equation}
two local time-pushes at different points fail to commute by a spatial
shift.

\emph{The aspiration.} Generators: framed-boundary filling swaps $(R,R')$
--- two distinct FK-legal fillings of the same boundary with matching
per-boundary-edge interior degrees (``framing''), the atomic moves that stay
exactly on the constraint surface. They compose partially (a groupoid), and
the desired structure lives in the relations: disjoint supports commute
(locality for free); the conjecture-shaped hope is that commutators of
overlapping moves generate the defect-transport (``shift'') moves --- a
discrete echo of $\{H,H\} \sim D$, in which the momentum sector is not added
but \emph{is} the commutator subalgebra. Status: kinematics built and
validated (framed boundaries, canonical forms, move detection by
enumeration); no explicit generator yet (the enumeration wall); the
algebraic layer unexplored. Structurally satisfying division of labor: in
ADM, $H$ is a term in the action while $D$ generates the gauge flow ---
\emph{penalties for $H$, moves for $D$} is arguably the correct
discretization, not a workaround.

%--------------------------------------------------------------------
\section{The frozen-subset sampler}
\label{sec:frozen}

Motivated by the shell-fixing idea (fix curvature on a spherical shell,
sample the rest --- boundary-condition and holographic experiments), a
subset-constrained sampler was designed and implemented in the D core.

\paragraph{Freezing tetrahedra does not fix curvature.} Curvature lives on
hinges, and an edge's deficit depends on \emph{all} incident tetrahedra,
including unfrozen ones outside the set; a move in the collar changes shell
deficits without touching frozen tetrahedra. To pin the curvature on a set
of hinges one must freeze the \emph{closed star} of those hinges.

\paragraph{Support containment makes vertex-freezing exact.} The
implementation freezes \emph{vertices}: the sampler rejects any move whose
support (bistellar ball $\leq 5$ vertices; hinge support $6$) contains a
frozen vertex. Since every facet a move adds or removes has all its vertices
in the support, vertex freezing preserves the frozen set's entire closed
star exactly --- facets \emph{and} all hinge degrees, hence all pinned
curvature; star-closure is built in via \texttt{freeze\_facets}. Hard
rejection by a fixed predicate restricts the chain to the sub-configuration
space and preserves detailed balance there. Storage is a byte array parallel
to the hot per-vertex degree array (memory-bound code; bit-stealing the
degree array was rejected as corruption risk across $\sim 25$ arithmetic
call sites for no measurable gain). Validation: D unittest (frozen closed
star exactly preserved over $4{,}000$ mixed MCMC steps) and an end-to-end
Python melt test ($300$ sweeps of full melt around a frozen shell; all $198$
star facets and $166$ hinge degrees bit-identical). Overhead $+2.0\%$ with
$17$ frozen vertices; feature-off fast path in the noise.

\paragraph{What it enables.} (1) \emph{Cavity sampling} --- freeze everything
except a small region; the sampler explores fillings of that exact framed
boundary through illegal intermediates; every FK-legal configuration visited
is a harvested block move, with realization path and relative weight ---
the move-enumeration wall bypassed by physics. (2) \emph{Gauss-law shell
test} --- freeze a star-closed shell around known curvature content and
measure the exterior response: the enclosed-charge experiment in real space,
bypassing box-size limits on $S(\mathbf k)$. (3) \emph{Measured Green's
functions} --- two shells differing by a localized curvature perturbation;
the difference of mean interior fields is the discrete Dirichlet Green's
function of the medium. (4) \emph{Entropy-area scaling} --- interior
ensemble entropy vs.\ boundary area, the holographically flavored
measurement. (5) Carrier confinement, defect pinning, interfaces, and gauge
anchoring.

%--------------------------------------------------------------------
\section{Penalties are not constraints}
\label{sec:penalties}

Three empirical strikes sharpened this theme into the program's central
lesson.

\paragraph{The bump ladder null.} Raising the demanded background charge
$5\times$ (bump $8\times10^{-4} \to 4\times10^{-3}$ at $\lambda=0.4$) left
the carrier density unchanged ($\langle n_{\rm ill}\rangle = 59, 63, 48$)
and the achieved mean degree pinned at $\approx 5.1043$: the pin simply goes
unsatisfied. Carrier density is thermal (set by the FK-penalty/entropy
balance), not charge-driven. Neutrality is \emph{requested}, never
\emph{enforced}: a real one-component-plasma background forces carriers into
existence; a soft pin only prices their absence.

\paragraph{Pin transmission is $\lambda$-dependent.} Where has the mean-degree
pin actually been satisfied? Three tiers: the certified disordered-fluid
seed library (satisfied exactly --- no crystal fighting back); the certified
$\lambda=1$ knot ensemble (achieved $5.104755$ vs.\ target $5.105025$:
$66\%$, with above- and below-initialized states agreeing to $10^{-6}$);
the $\lambda=0.4$ liquid ($\approx 15\%$, with near-zero marginal
susceptibility $d\langle\deg\rangle/d(\mathrm{bump})$). The soft pin is an
applied field; the crystal host at low $\lambda$ is curvature-incompressible.

\paragraph{Certification audit.} The $\lambda=0.4$ liquid ensemble narrowly
fails the formal gates ($5$ chains $\times$ $134$ frames):

\begin{center}
\begin{tabular}{lccl}
\toprule
observable & quantized split-$\hat R$ & ESS & verdict (gates $1.05$ / $100$) \\
\midrule
$n_{\rm illegal}$ & 1.062 & 83 & fails both \\
legal-vertex frac & 1.062 & 83 & fails both \\
$n_{\rm comp}$ & 1.037 & 125 & passes \\
$\langle \deg_e \rangle$ & 1.018 & 385 & passes \\
\bottomrule
\end{tabular}
\end{center}

\noindent plus all five chains one-sided (initialized from the crystal,
below in density). Remedy in flight: extend the five chains and add
above-side chains initialized from high-density $\lambda=1$ states, then
re-gate the pooled two-sided ensemble --- the protocol that certified the
knot study.

\paragraph{The general lesson.} No conservation law $\Rightarrow$ no sum
rules $\Rightarrow$ no Stillinger--Lovett \cite{stillinger-lovett}, at any
box size. Every long-range null collected in the liquid (ideal-gas carriers,
flat dielectric response, no SL dip, bump-independent density) is the same
fact viewed through different observables. Conversely, the frozen-vertex
infrastructure makes \emph{rejection-based hard constraints} cheap: a
microcanonical total-curvature window, or the budgeted-illegality sampler of
\S\ref{sec:s66}, are a few lines on a validated pattern.

%--------------------------------------------------------------------
\section{The six-web as an emergent gauge field; $S_{66}(\mathbf k)$}
\label{sec:s66}

The degree-6 web is the model's one genuinely topological structure, and its
legality condition maps onto emergent gauge theory
\cite{henley,castelnovo}:

\begin{center}
\begin{tabular}{ll}
\toprule
project object & gauge object \\
\midrule
six-edge web (closed line network) & field lines of an emergent flux \\
$\Nfk \in \{0,2,3,4\}$ at every vertex & lattice Gauss law (junctions allowed) \\
$\Nfk = 1$ vertex (line endpoint) & monopole (Gauss-law violation) \\
$z_{\rm leg}/c_{\rm imp}$ penalties & monopole \emph{mass}, not prohibition \\
knots / blinkers & thermally excited monopole plasma \\
\bottomrule
\end{tabular}
\end{center}

When the closure constraint is exact, constrained entropy alone generates a
Coulomb phase: algebraic dipolar correlations and pinch points in the
structure factor \cite{isakov,henley}. At finite violation density,
monopoles screen the emergent field with correlation width
$\xi^{-1} \propto \sqrt{n_{\rm mono}}$; our $\lambda=0.4$ liquid carries
$\sim 60$ illegal vertices in a $4^3$ box --- monopole spacing
$\lesssim 1$ cell, so the Coulomb physics is screened below the shortest
measurable wavelength. On $T^3$ the closed web additionally carries a
$\mathbb Z^3$ homology winding $W$ (net six-flux through the three
cross-sections), exactly conserved by local moves on the legal manifold ---
the clean superselection charge; monopoles let $W$ drift.

\paragraph{The observable.} Attach $s_e = 1$ to six-edges at midpoints
$\mathbf x_e$ (harmonic torus coordinates, minimum image). The scalar
channel is
\begin{equation}
S_{66}(\kv) = \frac{1}{N_6}
\Big\langle \Big| \sum_{e: \deg e = 6} e^{i \kv\cdot \mathbf x_e} \Big|^2
\Big\rangle,
\end{equation}
normalized against an \emph{occupation-shuffle null} (a random edge subset
of the same size on the same skeleton), which divides out the underlying
lattice. Since six-edges are unoriented segments with axes $\nh_e$, the
tensor channel uses the director (see \S\ref{sec:nematic}),
\begin{equation}
Q^{ab}_e = \nh^a_e \nh^b_e - \tfrac{\delta^{ab}}{3},
\qquad
Q^{ab}(\kv) = \sum_{e:\deg e = 6} Q^{ab}_e\, e^{i\kv\cdot\mathbf x_e},
\label{eq:director}
\end{equation}
helicity-decomposed about $\kh$ into transverse-traceless, vector, and
longitudinal parts, with $L(\kv) = \tfrac32 |\kh_a Q^{ab}(\kv) \kh_b|^2$.
Coarse-graining the web into a flux field $\mathbf B$, closure reads
$\nabla\!\cdot\!\mathbf B = \rho_{\rm mono} = 0$, i.e.\
$\kv\cdot\mathbf B(\kv) = 0$: longitudinal weight is forbidden mode by mode,
and for a maximally entropic web
$\langle B^a(\kv) B^b(-\kv)\rangle \propto \delta^{ab} - \kh^a\kh^b$
--- the pinch-point form \cite{youngblood-axe,isakov}. Monopoles restore
longitudinal weight over a width $\propto \sqrt{n_{\rm mono}}$.

\paragraph{Measured results} (single snapshots per state; ratios to
occupation-shuffle null; $L/T$ per $|\mathbf n|$ shell):

\begin{center}
\begin{tabular}{lccccc}
\toprule
state & $n_{\rm mono}$ & shell $1$--$2$ & shell $2$--$3$ & shell $3$--$5$ & pooled $L/T$, $k\in[1,2]$ \\
\midrule
crystal & 0 & \multicolumn{3}{c}{stealthy: $0.000$ sub-Bragg (all channels)} & --- \\
frozen $\lambda=1$ & 62 & 0.37 & 0.74 & 0.90 & $\mathbf{0.603 \pm 0.040}$ \\
liquid $\lambda=0.4$ & 61 & 0.53 & 1.34 & 0.77 & $0.844 \pm 0.064$ \\
$\lambda=0.35$ & 186 & 1.16 & 0.98 & 0.98 & $\approx 1$ \\
$\lambda=0.30$ web & 3642 & 0.91 & 1.15 & 0.75 & $\approx 1$ \\
\bottomrule
\end{tabular}
\end{center}

The pooled frozen-ensemble result, $L/T = 0.603 \pm 0.040$, is $10\sigma$
below unity, with the suppression strongest at smallest $k$ and dying with
monopole density exactly on the Coulomb-phase schedule: an emergent Gauss
law, measured, and monopole-screened at current densities. Notably the
frozen $\lambda=1$ ensemble suppresses better than the liquid at equal
monopole \emph{count}: violation \emph{activity} (flicker), not just census,
sets the screening. The agreed next step is the \emph{budgeted-illegality
sampler} --- reject any move pushing the global illegal count above a small
cap (a poor-man's worm algorithm \cite{prokofev-svistunov} on the
frozen-vertex rejection pattern; the $n_6$ potential already tracks the
count incrementally) --- driving $n_{\rm mono} \to 0$ while keeping the web
mobile, with the prediction $L/T \to 0$ and pinch points sharpening as the
budget tightens.

%--------------------------------------------------------------------
\section{Nematic vs.\ vector longitudinal channels}
\label{sec:nematic}

The director workaround in \eqref{eq:director} has a precise cost, worth
stating carefully because it changes what ``longitudinal suppression''
means.

\paragraph{Why ``nematic.''} A six-edge has an axis but no arrow; observables
must be invariant under $\nh_e \to -\nh_e$. The lowest surviving moment is
the quadrupole $Q_e = \nh_e\nh_e - \mathbf I/3$ --- precisely the order
parameter of a nematic liquid crystal (rod-like molecules with head--tail
symmetry) \cite{degennes}. The five components of the symmetric-traceless
$Q(\kv)$ split under rotations about $\kh$ into helicities $2+2+1$
(TT / vector / longitudinal); the helicity-0 part
$L = \tfrac32|\kh Q \kh|^2$ measures splay-like modulation of line density
along the line axis.

\paragraph{Why it is not the usual longitudinal channel.} The Gauss law is a
statement about a \emph{vector} flux: for oriented lines,
$\kv\cdot\mathbf J(\kv) = 0$ exactly, killing the vector-longitudinal
channel identically --- the sharp, quantitative signature
\cite{youngblood-axe}. Because $Q$ is quadratic in $\nh$, it inherits only a
blurred image of the constraint: line \emph{bending} leaks into
$\kh Q \kh$ even at exactly zero divergence. Hence frozen
$L/T = 0.60$ rather than $0$: an unknown fraction of the residual is
intrinsic nematic background, not monopole physics. The measurement
establishes the inequality and the monopole trend; the vector channel turns
it into an identity test.

\paragraph{The orientation obstruction, and its physical meaning.} Unit
fluxes $\sigma_e = \pm 1$ can only be divergence-free at even-valence
vertices; the six-web has Y-junctions (Z15, $\Nfk = 3$), where unit-flux
conservation is impossible. This is not a bookkeeping nuisance: TCP
disclinations compose non-abelianly ($\pi_1$ of the order-parameter space
\cite{mermin}), and the web is a string-net rather than a $U(1)$ loop gas.
The abelianized shadow is a \emph{real-valued} flow.

\paragraph{The exact-segment construction.} Two steps.

\emph{(i) Cycle-space projection.} Let $B$ be the head-minus-tail incidence
matrix of the six-web restricted to \emph{legal} vertex rows (divergence is
permitted only at edge-illegal, i.e.\ monopole, vertices). Draw random signs
$\sigma^0 \in \{\pm1\}^{E_6}$ and project onto the constraint:
\begin{equation}
\varphi = \arg\min \|\varphi - \sigma^0\|^2
\;\;\text{s.t.}\;\; B\varphi = 0
\qquad\Longleftrightarrow\qquad
BB^{\!\top}\mu = B\sigma^0,\quad \varphi = \sigma^0 - B^{\!\top}\mu,
\end{equation}
solved by sparse least squares (LSMR handles the rank deficiency of fully
closed components); averaging over draws of $\sigma^0$ is a Dirac-string
gauge average. Junction branches carry fractional flux.

\emph{(ii) Exact segment kernel.} Fourier the flux as a line integral along
each segment rather than a midpoint point charge. For an edge from
$\mathbf x_0$ along $\mathbf d$:
\begin{equation}
\int_0^1 \mathbf d\; e^{i\kv\cdot(\mathbf x_0 + t\mathbf d)}\,dt
= \mathbf d\, e^{i\kv\cdot\mathbf x_0}\,
e^{i\theta/2}\,\frac{\sin(\theta/2)}{\theta/2},
\qquad \theta = \kv\cdot\mathbf d,
\label{eq:kernel}
\end{equation}
so that the longitudinal projection \emph{telescopes}: dotting $\kv$ into
\eqref{eq:kernel} yields a pure endpoint difference
$\tfrac1i\!\left(e^{i\kv\cdot\mathbf x_{\rm head}} -
e^{i\kv\cdot\mathbf x_{\rm tail}}\right)$, and summing over edges,
\begin{equation}
\kv\cdot\mathbf J(\kv)
= \frac1i \sum_v D_v\, e^{i\kv\cdot\mathbf x_v},
\qquad D_v = (\text{net divergence of }\varphi\text{ at } v).
\label{eq:telescope}
\end{equation}
Every legal vertex drops out identically (commensurate $\kv$ on the torus
makes wrap-around phases exactly $1$), leaving
\begin{equation}
\boxed{\;
L_{\rm vec}(\kv) \;=\; |\kh\cdot\mathbf J(\kv)|^2
\;=\; \frac{S_{\rm mono}(\kv)}{k^2}\;,
\qquad
S_{\rm mono}(\kv) = \Big|\sum_v D_v e^{i\kv\cdot\mathbf x_v}\Big|^2
}
\label{eq:identity}
\end{equation}
--- the longitudinal power \emph{is} the monopole-charge structure factor
over $k^2$: identically zero on a closed web, and a direct census-with-
positions of Gauss-law violations otherwise. As a bonus, the small-$k$
behavior of $S_{\rm mono}$ tests whether the \emph{monopole plasma itself}
screens: $S_{\rm mono} \to \sum_v D_v^2$ (flat) for uncorrelated monopoles,
vs.\ $S_{\rm mono} \sim k^2/\kappa^2 \to 0$ for a Stillinger--Lovett-obeying
emergent-charge plasma --- tying the hyperuniformity and gauge threads
together in one observable.

\paragraph{Implementation and first light.} Implemented as
\texttt{scripts/defect\_dynamics/web\_vector.py} (cycle-space projection via
LSMR; exact kernel \eqref{eq:kernel}; per-state numerical verification of
\eqref{eq:telescope}). First run across the monopole ladder
(identity \eqref{eq:telescope} verified per state to absolute error
$\sim\!10^{-10}$ on channel powers of order $10^{10}$; residual legal-vertex
divergence $\sim\!10^{-13}$):

\begin{center}
\begin{tabular}{lcccccc}
\toprule
state & $n_{\rm mono}$ & $\sum_v D_v^2$ & $L/T$ (shell $1$--$2$)
& \multicolumn{3}{c}{$S_{\rm mono}/\sum D_v^2$ by shell ($1$--$2$, $2$--$3$, $3$--$5$)} \\
\midrule
crystal & 0 & 0 & $3\times10^{-26}$ & \multicolumn{3}{c}{--- (no sources)} \\
frozen $\lambda=1$ & 62 & 172 & $1.5\times10^{-2}$ & 0.12 & 0.17 & 0.31 \\
liquid $\lambda=0.4$ & 61 & 173 & $1.2\times10^{-2}$ & 0.13 & 0.23 & 0.45 \\
$\lambda=0.35$ & 186 & 422 & $6.4\times10^{-2}$ & 0.21 & 0.36 & 0.55 \\
$\lambda=0.30$ web & 3642 & 8304 & $4.1\times10^{-1}$ & 0.10 & 0.26 & 0.45 \\
\bottomrule
\end{tabular}
\end{center}

\noindent The crystal's closed web gives $L/T$ at machine zero --- the
nematic $0.60$ upgraded to an exact null. In every monopole-bearing state
the longitudinal weight is purely monopole-sourced, and
$S_{\rm mono}(\kv)$ is suppressed at small $k$ to $10$--$20\%$ of its
uncorrelated (Poisson) level $\sum_v D_v^2$, recovering toward it at large
$k$: the emergent monopole charges are themselves strongly anticorrelated
at long wavelength --- screening \emph{among} the violations, a
Stillinger--Lovett-type organization of the monopole plasma visible in the
very observable that certifies the Gauss law. The budget-dependence of both
numbers is the budgeted-illegality campaign's target.

%--------------------------------------------------------------------
\section{Outlook}
\label{sec:outlook}

\begin{enumerate}
\item \emph{Certification} of the $\lambda=0.4$ liquid (two-sided campaign
running), then of the $\lambda_{\rm EDQ}=1.2$ two-term gas.
\item \emph{Reproduce the liquid-state results under the pure two-term
action} (channel hierarchy, $L/T$) to confirm nothing depended secretly on
the $n_6$ scaffolding.
\item \emph{Budgeted-illegality sampler}: cap the global illegal count via
the validated rejection pattern; drive the monopole budget down; watch
$L_{\rm vec} \to 0$ via \eqref{eq:identity} and the pinch points sharpen;
then $W$-sector physics on the legal manifold.
\item \emph{Block-move harvest by cavity sampling} (frozen-subset sampler),
targeting the first explicit FK$\to$FK generators and their commutators ---
the momentum-sector program.
\item \emph{C15 host} for host-independence and curved-background controls.
\end{enumerate}

%--------------------------------------------------------------------
\begin{thebibliography}{99}

\bibitem{adm} R.~Arnowitt, S.~Deser, C.~W.~Misner,
\emph{The dynamics of general relativity}, in \emph{Gravitation: An
Introduction to Current Research}, ed.\ L.~Witten (Wiley, 1962);
reprinted Gen.\ Rel.\ Grav.\ \textbf{40}, 1997 (2008).

\bibitem{regge} T.~Regge,
\emph{General relativity without coordinates},
Nuovo Cimento \textbf{19}, 558 (1961).

\bibitem{hamber-williams} H.~W.~Hamber, R.~M.~Williams,
\emph{Higher derivative quantum gravity on a simplicial lattice},
Nucl.\ Phys.\ B \textbf{248}, 392 (1984).

\bibitem{ajl} J.~Ambj{\o}rn, J.~Jurkiewicz, R.~Loll,
\emph{Nonperturbative quantum gravity},
Phys.\ Rep.\ \textbf{519}, 127 (2012).

\bibitem{pachner} U.~Pachner,
\emph{P.L.\ homeomorphic manifolds are equivalent by elementary shellings},
Eur.\ J.\ Combin.\ \textbf{12}, 129 (1991).

\bibitem{sorkin} R.~Sorkin,
\emph{Time-evolution problem in Regge calculus},
Phys.\ Rev.\ D \textbf{12}, 385 (1975).

\bibitem{dittrich-hohn} B.~Dittrich, P.~A.~H\"ohn,
\emph{Canonical simplicial gravity},
Class.\ Quantum Grav.\ \textbf{29}, 115009 (2012).

\bibitem{bahr-dittrich} B.~Bahr, B.~Dittrich,
\emph{(Broken) gauge symmetries and constraints in Regge calculus},
Class.\ Quantum Grav.\ \textbf{26}, 225011 (2009).

\bibitem{thiemann} T.~Thiemann,
\emph{Quantum spin dynamics (QSD)},
Class.\ Quantum Grav.\ \textbf{15}, 839 (1998).

\bibitem{frank-kasper} F.~C.~Frank, J.~S.~Kasper,
\emph{Complex alloy structures regarded as sphere packings I, II},
Acta Cryst.\ \textbf{11}, 184 (1958); \textbf{12}, 483 (1959).

\bibitem{nelson} D.~R.~Nelson,
\emph{Order, frustration, and defects in liquids and glasses},
Phys.\ Rev.\ B \textbf{28}, 5515 (1983).

\bibitem{sadoc-mosseri} J.-F.~Sadoc, R.~Mosseri,
\emph{Geometrical Frustration} (Cambridge University Press, 1999).

\bibitem{grunbaum-motzkin} B.~Gr\"unbaum, T.~S.~Motzkin,
\emph{The number of hexagons and the simplicity of geodesics on certain
polyhedra}, Can.\ J.\ Math.\ \textbf{15}, 744 (1963).

\bibitem{mermin} N.~D.~Mermin,
\emph{The topological theory of defects in ordered media},
Rev.\ Mod.\ Phys.\ \textbf{51}, 591 (1979).

\bibitem{henley} C.~L.~Henley,
\emph{The ``Coulomb phase'' in frustrated systems},
Annu.\ Rev.\ Condens.\ Matter Phys.\ \textbf{1}, 179 (2010).

\bibitem{castelnovo} C.~Castelnovo, R.~Moessner, S.~L.~Sondhi,
\emph{Magnetic monopoles in spin ice},
Nature \textbf{451}, 42 (2008).

\bibitem{isakov} S.~V.~Isakov, K.~Gregor, R.~Moessner, S.~L.~Sondhi,
\emph{Dipolar spin correlations in classical pyrochlore magnets},
Phys.\ Rev.\ Lett.\ \textbf{93}, 167204 (2004).

\bibitem{youngblood-axe} R.~W.~Youngblood, J.~D.~Axe,
\emph{Polarization fluctuations in ferroelectric models},
Phys.\ Rev.\ B \textbf{23}, 232 (1981).

\bibitem{prokofev-svistunov} N.~Prokof'ev, B.~Svistunov,
\emph{Worm algorithms for classical statistical models},
Phys.\ Rev.\ Lett.\ \textbf{87}, 160601 (2001).

\bibitem{stillinger-lovett} F.~H.~Stillinger, R.~Lovett,
\emph{General restriction on the distribution of ions in electrolytes},
J.\ Chem.\ Phys.\ \textbf{49}, 1991 (1968).

\bibitem{torquato} S.~Torquato,
\emph{Hyperuniform states of matter},
Phys.\ Rep.\ \textbf{745}, 1 (2018).

\bibitem{degennes} P.-G.~de Gennes, J.~Prost,
\emph{The Physics of Liquid Crystals}, 2nd ed.\ (Oxford University Press,
1993).

\end{thebibliography}

\end{document}
